### 2016-03-17 doc/thesis/intro.tex
Mostly writing

\section{Phylogenetics and phylodynamics}

@@

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

\begin{figure}[ht]
  \centering
  \label{fig:speciestree}
  \includegraphics{speciestree}
  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
    {Illustration of a rooted, ultrametric, time-scaled phylogeny. The tips of
      the tree, which represent extant taxa, are placed at the present day on
      the time axis. Internal nodes, representing extinct common ancestors to
      the extant taxa, fall in the past. The topology of the tree indicates
      that cats and dogs are the most closely related pair of species, whereas
      fish is most distantly related to any other node in the tree.}
\end{figure}

\subsection{Transmission trees and epidemiology}

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiolgical process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with a root corresponding to the index case, and branching times indicating
times of transmission. The internal nodes may be labelled with
the donor of the transmission pair, if this is known. The tips of the tree,
rather than being fixed at the present day, are placed at the time at which the
individual was removed from the epidemic, such as by death, recovery,
isolation, behaviour change, or migration. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous}
taxa~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} taxa found in the phylogenies of macro-organisms.  A
transmission tree is illustrated in Figure~\ref{fig:virustrees}.

\begin{figure}
    \centering
    \includegraphics{virustrees}
    \label{fig:virustrees}
    \caption[Illustration of a transmission tree]{
        A transmission tree representing an epidemic's progress through
        four individuals. The index case, patient $a$, transmitted the
        virus to patients $b$ and $c$; patient $c$ went on to infect
        patient $d$. 
    }
\end{figure}

Since transmission trees are essentially a detailed record of an epidemic's
progress, they contain substantial epidemiological information. As a basic
example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
lineages in a phylogeny against time, can be used to quantify the incidence of
new infections over the course of an epidemic~\autocite{holmes1995revealing}.
Many more diverse epidemiological parameters have been investigated using
transmission trees, such as the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple}. However, in all
but the most well-studied of epidemics, this is not possible to obtain through
traditional epidemiological methods. The time and effort to conduct detailed
interviews and contact tracing of a sufficient number of infected individuals
is usually prohibitive. Even when the resources for such methods are available,
patients may not always recall whom they contacted and when, especially in the
case of airborne transmission. Consequently, the transmission tree must be
estimated using other methods. Most commonly, this is done by exploiting the
relationship between transmission trees and viral phylogenies.

\subsection{Estimating transmission trees from viral phylogenies}

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. \comment{Is ``strain'' the best word here?} In phylodynamics, we
often consider \defn{inter-host} phylogenies, which relate one viral genotype
from each host in a population. The tips of the tree are labelled with both the
virus and its host. The crux of
\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
processes generating these two types of tree - epidemiological processes, for
transmission trees, and evolutionary processes, for viral phylogenies - occur
on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
result, there is a close relationship between the two trees. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual causes the
extinction of their viral lineage. Due to these relationships, the topology of
the viral phylogeny is often used as a proxy for the topology of the
transmission tree. However, there are several complications and caveats which
must be kept in mind when estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction produce unrooted trees whose branch
lengths measure genetic distance in units of expected substitutions per site,
whereas transmission trees are rooted with branches measuring calendar time. We
generally assume that sampling a virus from the individual also corresponds to
their removal from the transmission tree\comment{needs a reference}, so the
positions of the tips in time are fixed. Therefore, to transform a viral
phylogeny into an estimated transmission tree, we must find a root and an
assignment of branch lengths such that the tips are placed at their proscribed
times. Of course, there are an infinite number of possible combinations of root
and branch lengths which result in the correct placement of the tips, so we
would also like to find values which are the most ``reasonable'' given our
data. By reasonable, we mean that the variation in \defn{evolutionary rate},
which is the ratio of a branch's length in genetic distance to its length in
calendar time, is low across the tree. While there is some variation among
hosts, due to immunological and other factors, we generally expect to observe
globally similar evolutionary rates. Methods for time-scaling a phylogeny
include root-to-tip regression~\autocite{shankarappa1999consistent,
korber2000timing, drummond2003inference}, which we apply in this work, and
least-square dating~\autocite{to2015fast}. Both of these methods can be used to
root the tree, by simply trying all possible root positions and choosing the
one which minimizes a loss function ($1-R^2$, or root-mean-square error, for
root-to-tip regression; sum of squared errors for least-square dating).
Alternatively, the tree may be rooted with an outgroup~\autocite{li1988rates}
before time-scaling.

@@

A final caveat is that the viral phylogeny itself is not known with certainty,
so it must also be estimated from genetic data. Phylogenetic inference is a
complex topic which we will not discuss in detail here (see e.g.
\autocite{nei2000molecular} for a full review). Most modern analyses use
model-based methods, which simultaneously estimate the phylogeny with branch
lengths and the parameters of a model of evolution. Although they usually work
well in practice \comment{needs a reference}, the estimated topology can vary
based on the model used and, in the case of Bayesian analysis, the priors.  In
addition, intra-host viral populations are genetically heterogeneous, so
choosing a single representative genotype per host is necessarily imprecise.
One can use either the genotype of a specific virion sampled from the host, or
a synthetic genotype, such as a consensus or reconstructed ancestral sequence.

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) \comment{are
tip dates part of the prior?} can be included in the analysis, and the
so-called nuisance parameters of the molecular clock model can be marginalized
out. Software packages for performing these analyses include
BEAST~\autocite{bouckaert2014beast} and MrBayes~\autocite{ronquist2012mrbayes}.

Several other authors have developed methods tailor-made for inferring
transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
of the two-step approach which allows transmissions to occur anywhere along the
branches of a transmission tree, rather than being constrained to the branching
points in the viral phylogeny. The method requires sampling of every infected
individual, although the authors indicate that it could be extended to relax
this assumption. \textcite{cottam2008integrating} describe a likelihood-based
method which enumerates all transmission trees consistent with an established
phylogeny, assigning each a likelihood based on other epidemiological data. 
This approach is novel in its integration of data from multiple sources,
however because it enumerates a large portion of the tree space, it is unlikely
to scale to larger epidemics. A different approach is undertaken by
\textcite{jombart2011reconstructing}, who describe a method to build
transmission trees directly from sequence data, contingent on the common
ancestors also being sampled. This makes the method attractive for
slow-evolving pathogens, but less practical for viral outbreaks where samples
from common ancestors are unlikely to be available.

@@

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often an estimated transmission tree~\autocite{pybus2009evolutionary,
volz2013viral}. Phylogenies are complex objects, and it is not immediately
obvious how to extract useful information from them with respect to fitting a
parameter. Standard statistical methods built for numeric data cannot be
applied directly - for example, one cannot perform a regression of a parameter
of interest against a phylogeny. Therefore, before we discuss exactly what
phylodynamics can tell us about epidemics, and how it has been applied in the
past, we will review some of the methods for quantifying the shapes of
phylogenies and their similarity to each other.

The shape of a phylogeny consists of two components: the topology, and the
distribution of branch lengths~\autocite{mooers1997inferring}. Many tree
summary statistics have been developed to assign numerical values to
phylogenies based on their shapes. One of the most widely used is Sackin's
index~\autocite{shao1990tree}, which measures the imbalance or asymmetry in a
rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
branches between that tip and the root. The unnormalized Sackin's index is
defined as the sum of all $N_i$. It is called unnormalized because it does not
account for the number of tips in the tree. Among two trees having the same
number of tips, the least-balanced tree will have the highest Sackin's index.
However, among two equally balanced trees, the larger tree will have a higher
Sackin's index. This makes it challenging to compare balances among trees of
different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
the expected value of Sackin' index under the Yule
model~\autocite{yule1925mathematical}. Dividing by this expected value
normalizes Sackin's index, so that it can be used to compare trees of different
sizes.

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure associates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1. In
the context of infectious diseases, the \gls{ltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
kernel is a \defn{similarity measure}, rather than a distance measure, as it is
maximized when the two trees being compared are the same. The basis of the tree
kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibitive. In
the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
to fit epidemiological models to observed trees.

\subsection{Applications of phylodynamics}
\label{subsubsec:appphylo}

\section{Contact networks}
\label{subsec:contactnet}

\subsection{Overview}

Epidemics spread through populations of hosts through \defn{contacts} between
those hosts. The definition of contact depends on the mode of transmission of
the pathogen in question. For an airborne pathogen like influenza, a contact
may be simple physical proximity, while for a sexually transmitted pathogen
like HIV, contacts would be sexual partnerships. A \defn{contact network} is a
graphical representation of a host population and the contacts among its
members. The \defn{nodes} in the network represent hosts, and \defn{edges}
represent contacts between them.

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, a blood-borne infection spread through
transfusions would use undirected edges, since the donor has no chance of
transmitting to the recipient. Directed edges are also useful when the
transmission risk is not equal between the hosts, such as with HIV
transmission, where acting as the receptive partner carries a higher risk of
infection than acting as the insertive partner. In this case, a contact could
be represented by two directed edges, one in each direction between the two
hosts, with the edges annotated by what kind of risk they imply. In fact, it is
possible to represent an undirected edge by two symmetric directed edges. For
this reason, we consider only contact networks with directed edges in the
sequel. A directed contact network is shown in Figure \ref{fig:contactnet}
(left).

The path an epidemic takes through a contact network determines the topology of
the transmission tree relating the infected hosts. The initially infected node
who introduces the epidemic becomes the root of the tree. Each time a
transmission occurs, the lineage corresponding to the donor host in the tree
splits into two, representing the recipient lineage and the continuation of the
donor lineage. This correspondence is illustrated in figure
\ref{fig:contactnet}. It's important to note that, although the order and
timing of transmissions determines the tree topology uniquely, the converse
does not hold. That is, there are generally several orders of infection which
could lead to the same topology, since the labels on the internal nodes of the
tree are not available to the researcher.

\begin{figure}[ht]
  \centering
  \label{fig:contactnet}
  \includegraphics{contactnet}
  \caption[Illustration of a epidemic spread over a contact network]
  {Illustration of epidemic spread over a contact network. On the left, a
   contact network with five hosts, labelled A through E. Directed edges
   indicate six symmetric contacts among the hosts. Coloured, bolded edges
   represent transmissions. The epidemic began with node A, who transmitted to
   nodes B and C. Node C further transmitted to node D, and node E was not
   infected. On the right, the transmission tree topology corresponding to this
   scenario. Note that individual B was sampled before the other infected
   individuals, resulting in a tree with heterochronous tips.}
\end{figure}

\subsection{Models for contact networks}
\label{subsubsec:generative}

Throughout this section, the variable $n$ will be used to refer to the number
of nodes in the network.

An \defn{Erd\H{o}s-R\'enyi} (ER) network~\autocite{erdos1960evolution}, also
known as a Bernoulli network, was the first network model described. It has a
single parameter, $p$, which is the probability that any particular edge is
present in the network. To construct an ER network, one simply chooses
$p \times \binom{n}{2}$ distinct pairs of nodes and draws an edge between each
pair. The average degree of a node in an ER network is equal to $p \times n$.

>>>>

\section{Phylogenetics and phylodynamics}
\label{sec:phylo}

@@

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
\cref{fig:speciestree}.

\begin{figure}[ht]
  \centering
  \includegraphics{speciestree.pdf}
  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
    {Illustration of a rooted, ultrametric, time-scaled phylogeny. The tips of
      the tree, which represent extant taxa, are placed at the present day on
      the time axis. Internal nodes, representing extinct common ancestors to
      the extant taxa, fall in the past. The topology of the tree indicates
      that cats and dogs are the most closely related pair of species, whereas
      fish is most distantly related to any other node in the tree.}
  \label{fig:speciestree}
\end{figure}

\subsection{Transmission trees}

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with a root corresponding to the index case, and branching times indicating
times of transmission. The internal nodes may be labelled with
the donor of the transmission pair, if this is known. The tips of the tree,
rather than being fixed at the present day, are placed at the time at which the
individual was removed from the epidemic, such as by death, recovery,
isolation, behaviour change, or migration. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous}
taxa~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} taxa found in the phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

In a transmission tree, each sampled individual appears as a tip when they are
removed from the infected population. In addition, they appear as an internal
node each time they transmit to an individual who is either eventually sampled,
or who transmits directly or indirectly to a sampled individual. This is not
the same as the transmission \emph{network}, in which each infected individual
appears exactly once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occured,
following~\autocite{welch2011statistical, keeling2005networks}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{contactnet.pdf}
    \caption[Illustration of a contact network and transmission tree]{
      Illustration of epidemic spread over a contact network, and the
      corresponding transmission tree. (Left) A contact network with five
      hosts, labelled $a$ through $e$. Thick shaded edges indicate symmetric
      contacts among the hosts. The transmission network is indicated by
      coloured arrows. The epidemic began with node $a$, who transmitted to
      nodes $b$ and $c$. Node $c$ further transmitted to node $d$. Node $e$ was
      not infected. (Right) The transmission tree corresponding to this
      scenario, with a timeline of transmission and removal times.
    }
    \label{fig:contactnet}
\end{figure}

Since transmission trees are essentially a detailed record of an epidemic's
progress, they contain substantial epidemiological information. As a basic
example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
lineages in a phylogeny against time, can be used to quantify the incidence of
new infections over the course of an epidemic~\autocite{holmes1995revealing}.
Many more diverse epidemiological parameters have been investigated using
transmission trees, such as the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple}. However, in all
but the most well-studied of epidemics, this is not possible to obtain through
traditional epidemiological methods~\autocite{welch2011statistical}. The time
and effort to conduct detailed interviews and contact tracing of a sufficient
number of infected individuals is usually prohibitive. Even when the resources
for such methods are available, patients may not always recall whom they
contacted and when, especially in the case of airborne transmission.
Consequently, the transmission tree must be estimated using other methods. Most
commonly, this is done by exploiting the relationship between transmission
trees and viral phylogenies~\autocite{volz2013viral}.

\subsection{Relationship between transmission trees and viral phylogenies}

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. In phylodynamics, we often consider \defn{inter-host}
phylogenies, which relate one viral genotype from each host in a population.
The tips of the tree are labelled with both the virus and its host. The crux of
\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
processes generating these two types of tree - epidemiological processes, for
transmission trees, and evolutionary processes, for viral phylogenies - occur
on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
result, there is a close relationship between the two trees. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual causes the
extinction of their viral lineage. Due to these relationships, the topology of
the viral phylogeny is often used as a proxy for the topology of the
transmission tree. However, there are several complications and caveats which
must be kept in mind when estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction produce unrooted trees whose branch
lengths measure genetic distance in units of expected substitutions per site,
whereas transmission trees are rooted with branches measuring calendar time. We
generally assume that sampling a virus from the individual also corresponds to
their removal from the transmission tree, so the positions of the tips in time
are fixed. Therefore, to transform a viral phylogeny into an estimated
transmission tree, we must find a root and an assignment of branch lengths such
that the tips are placed at their proscribed times. Of course, there are an
infinite number of possible combinations of root and branch lengths which
result in the correct placement of the tips, so we would also like to find
values which are the most ``reasonable'' given our data. By reasonable, we mean
that the variation in \defn{evolutionary rate}, which is the ratio of a
branch's length in genetic distance to its length in calendar time, is low
across the tree. While there is some variation among hosts, due to
immunological and other factors, we generally expect to observe globally
similar evolutionary rates. Methods for time-scaling a phylogeny include
root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Both of these methods can be used to root the
tree, by simply trying all possible root positions and choosing the one which
minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
regression; sum of squared errors for least-square dating). Alternatively, the
tree may be rooted with an outgroup~\autocite{li1988rates} before time-scaling.

@@

A final caveat is that the viral phylogeny itself is not known with certainty,
so it must also be estimated from genetic data. Phylogenetic inference is a
complex topic which we will not discuss in detail here (see e.g.
\autocite{nei2000molecular} for a full review). Most modern analyses use
model-based methods, which simultaneously estimate the phylogeny with branch
lengths and the parameters of a model of evolution. Although they usually work
well in practice, the estimated topology can vary based on the model used and,
in the case of Bayesian analysis, the priors.  In addition, intra-host viral
populations are genetically heterogeneous, so choosing a single representative
genotype per host is necessarily imprecise. One can use either the genotype of
a specific virion sampled from the host, or a synthetic genotype, such as a
consensus or reconstructed ancestral sequence.

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) 
%\comment{are tip dates part of the prior?} 
can be included in the analysis, and the so-called nuisance parameters of the
molecular clock model can be marginalized out. Software packages for performing
these analyses include BEAST~\autocite{bouckaert2014beast} and
MrBayes~\autocite{ronquist2012mrbayes}.

Several other authors have developed methods tailor-made for inferring
transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
of the two-step approach which allows transmissions to occur anywhere along the
branches of a transmission tree, rather than being constrained to the branching
points in the viral phylogeny. The method requires sampling of every infected
individual, although the authors indicate that it could be extended to relax
this assumption. \textcite{cottam2008integrating} describe a likelihood-based
method which enumerates all transmission trees consistent with an established
phylogeny, assigning each a likelihood based on other epidemiological data. 
This approach is novel in its integration of data from multiple sources,
however because it enumerates a large portion of the tree space, it is unlikely
to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint 
likelihood function integrating temporal, geographic, and genetic observations, 
and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
likelihood function. Their approach can handle missing data and produces high
resolution transmission trees when multiple types of data are available. A
different approach is undertaken by \textcite{jombart2011reconstructing}, who
describe a method to build transmission trees directly from sequence data,
contingent on the common ancestors also being sampled. This makes the method
attractive for slow-evolving pathogens, but less practical for viral outbreaks
where samples from common ancestors are unlikely to be available.

@@

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often an estimated transmission tree~\autocite{pybus2009evolutionary,
volz2013viral}. What is informative about a phylogeny, beyond the demographic
characteristics of the individuals it relates, is its \defn{shape}. The shape
of a phylogeny has two components: the topology, and the distribution of branch
lengths~\autocite{mooers1997inferring}. Methods of quantifying tree shape fall
into two categories: summary statistics, and pairwise measures.

Summary statistics assign a numeric value to each individual tree. One of the
most widely used is Sackin's index~\autocite{shao1990tree}, which measures the
imbalance or asymmetry in a rooted tree. For the $i$th tip of the tree, we
define $N_i$ to be the number of branches between that tip and the root. The
unnormalized Sackin's index is defined as the sum of all $N_i$. It is called
unnormalized because it does not account for the number of tips in the tree.
Among two trees having the same number of tips, the least-balanced tree will
have the highest Sackin's index. However, among two equally balanced trees, the
larger tree will have a higher Sackin's index. This makes it challenging to
compare balances among trees of different sizes. To correct this,
\textcite{kirkpatrick1993searching} derive the expected value of Sackin' index
under the Yule model~\autocite{yule1925mathematical}. Dividing by this expected
value normalizes Sackin's index, so that it can be used to compare trees of
different sizes.

Rather than assigning numbers to individual trees, pairwise measures associate
a numeric value to each \textit{pair} of trees, indicating how different the
trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1. In
the context of infectious diseases, the \gls{ltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. As opposed to the \gls{nltt},
the tree kernel is maximized when the two trees being compared are the same.
The basis of the tree kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibitive.
In the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
will be discussed further in \cref{sec:abc}.

\section{Contact networks}
\label{sec:contactnet}

\subsection{Overview}
\label{subsec:netoverview}

Epidemics spread through populations of hosts through \defn{contacts} between
those hosts. The definition of contact depends on the mode of transmission of
the pathogen in question. For an airborne pathogen like influenza, a contact
may be simple physical proximity, while for \gls{HIV}, contact could be via
unprotected sexual relations or blood-to-blood contact (such as through needle
sharing). A \defn{contact network} is a graphical representation of a host
population and the contacts among its members~\autocite{klovdahl1985social,
morris1993epidemiology, keeling2005networks}. The \defn{nodes} in the network
represent hosts, and \defn{edges} or \defn{links} represent contacts between
them. A contact network is shown in \cref{fig:contactnet} (left).

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, an infection which may be spread through
blood-to-blood contact through transfusions transfusions would use undirected
edges, since the donor has no chance of transmitting to the recipient. Directed
edges are also useful when the transmission risk is not equal between the
hosts, such as with HIV transmission, where acting as the receptive partner
carries a higher risk of infection than acting as the insertive partner. In
this case, a contact could be represented by two directed edges, one in each
direction between the two hosts, with the edges annotated by what kind of risk
they imply~\autocite{wasserman1994social}. An undirected contact network is
equivalent to a directed network where each contact is represented by two
symmetric directed edges. The \defn{degree} of a node in the network is how
many contacts it has. In directed networks, we may make the distinction between
\defn{out-degree} and \defn{in-degree}, which count respectively the number
incoming and outgoing edges. The set of edges attached to a node are referred
to as its \defn{incident} edges.

The path an epidemic takes through a contact network defines the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). The
transmission network determines the topology of the transmission tree relating
the infected hosts. The initially infected node who introduces the epidemic
into the network, also known as the \defn{index case}, becomes the root of the
tree. Each time a transmission occurs, the lineage corresponding to the donor
host in the tree splits into two, representing the recipient lineage and the
continuation of the donor lineage. \Cref{fig:contactnet} illustrates this
correspondence. It's important to note that, although the order and timing of
transmissions determines the tree topology uniquely, the converse does not
hold. That is, for any given topology, there are in general many transmission
networks which would lead to that topology. In other words, it impossible to
distinguish who transmitted to whom from a transmission tree
alone~\autocite{bernard2007hiv}.

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical}. For example, epidemics are
expected to spread more rapidly in networks having the ``small world''
property, where the average path length between two nodes in the network is
relatively low~\autocite{watts1998collective}. Some sexually transmitted
infections would not be expected to survive in a homogeneously mixed
population, but their long-term persistance can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
irradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks}. Several less resource-intensive alternatives have been
explored. Some work focuses on the phylogenetic network, detecting patterns
such as clustering. However, the phylogenetic network may show little to no
agreement with a contact data obtained through epidemiological
methods~\autocite{yirrell1998molecular, resik2007limitations}. Some
studies~\autocite{brown2011transmission} have focused on parameters of the
transmission network, rather than the more complete contact network, because it
is possible to derive an estimate of the transmission network phylogenetically.

Rather than considering entire social networks, some epidemiological research
has focused on properties which can be investigated without extensive contact
tracing. The most common of these properties is the \defn{degree distribution}
of the network, which denotes the probability that a node has any given number
of links. Degree distributions for sexual networks can be estimated by simply
asking each person how many contacts they had in, say, the last
year~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
